---
title: "CIMTD"
author: "Leila Bautista and Brendan Dwyer"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Canadian International Merchandise Trade Database (CIMTD)

This RMarkdown provides chunks of R code to load, filter and combine data from Open Government [website](https://search.open.canada.ca/opendata/?sort=metadata_modified+desc&search_text=cimt&page=1). The trade data is used by the Ministry of Agriculture and Food's Sector Insights and Corporate Initiatives (SICI) Unit to produce annual Sector Tables, Trade Dashboards and other information products.

Contact [alstats\@gov.bc.ca](mailto:alstats@gov.bc.ca) for any questions.

# Set-up & Library

```{r Set-up, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory

knitr::opts_knit$set(root.dir = "//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")
```

```{r Library}

# Load library
library(pacman)
p_load(dplyr,
       data.table,
       tidyr,
       lubridate,
       tictoc,
       splitstackshape,
       stringr)
```

# Load and read latest data

```{r Export Latest}
tic()

# Step 1: Create a temporary file
temp_file <- tempfile()

# Step 2: Fetch the online file using download.file()
# Replace 'url_of_the_file' with the actual URL of the file you want to download
url <- "https://www150.statcan.gc.ca/n1/pub/71-607-x/2021004/zip/CIMT-CICM_Dom_Exp_2024.zip"
download.file(url, temp_file)

# Step 3: Extract the file from the temp file using unz() command
# Replace 'file_inside_zip' with the actual name of the file inside the ZIP archive
file_inside_zip <- "CIMT-CICM_Dom_Exp_2024/ODPFN016_202405N.csv"
Export <- read.csv(unz(temp_file, file_inside_zip))

# Step 4: Remove the temp file using unlink()
unlink(temp_file)

# Load the list of agriculture, seafood and food manufacturing HS codes 
Commodity_REF <- fread("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/HS8X 2022 AAFC REF.csv",
                       colClasses = c(HS8_Code = "character"))

# Merge the file with the Commodity REF to fileter for agriculture, seafood and food manufacturing HS codes only
Export2024 <- Export %>% 
  rename (Date = `YearMonth.AnnéeMois`,
          HS8_Code = `HS8`,
          Coutnry = `Country.Pays`,
          State = `State.État`,
          Value = `Value.Valeur`,
          Quantity = `Quantity.Quantité`,
          UOM = `Unit.of.Measure.Unité.de.Mesure`
  ) %>%
  merge(Commodity_REF,by="HS8_Code") %>%
  select(1:8)

# reformat the date column
Export2024$Date <- ym(Export2024$Date)

# Set the working directory to the destination folder
setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")

# Write the csv file 
fwrite(Export2024, "Export2024.csv")  

# to write code: bind with Trade FULL


toc()
```

```{r Import Latest}
# To be completed
```

# Load and read 1988-2002 data

This chunk is to download from the Open Government website. Only Import data from 1988-2002 needed to be down because the export data 1988-2023 was downloaded through a past version of this R script.

```{r Import Historic}

tic()

# Load the list of agriculture, seafood and food manufacturing HS codes 

Commodity_REF <- fread("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/HS10M 2022 AAFC REF.csv",
                       colClasses = c(HS10_Code = "character"))


setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")


# Loop to load and write the data

for (i in 1988:2002) {
  
  # Paste year (1988-2023) to standardized file path to create file name

  Import_Filename <- paste0("ODPFN014_", i, "12C.csv", sep="")
  
  # Load each file name for all years (1988-2002)
  
  Import <- fread(Import_Filename,
                  colClasses = c(HS10 = "character"))
  
  # Rename variables to a useable format
  
  ImportYear <- Import %>%
    rename (Date = `YearMonth/AnnéeMois`,
            HS10_Code = `HS10`,
            Country = `Country/Pays`,
            State = `State/État`,
            Value = `Value/Valeur`,
            Quantity = `Quantity/Quantité`,
            UOM = `Unit of Measure/Unité de Mesure`
            ) %>%
    
    # Merge data with Commodity_REF to filter for agriculture, seafood and food manufacturing HS Codes only
    
    merge(Commodity_REF,by="HS10_Code") %>%
    
    # Drop all other variables other than 1:8
    
    select(1:8)
  
  Import_Outfile_Start <- paste("Import", i, sep="")
  
  Import_Outfile <- paste(Import_Outfile_Start,".csv", sep="")
  
  fwrite(ImportYear, Import_Outfile)
}

toc()
```

# Append all years into 1 file

The preceding code chunk has downloaded, unzipped and extracted all files from 1988 to the most recent year. The following chunks load and bind all years:

1.  Exports

2.  Imports

3.  Trade

## 1. Exports: Load and Bind All Years

The code chunk under this heading is an example of how to append all annual ***import*** data frames found in the [Annual](%22\sfp.idir.bcgov\S104\S04004\Statistics%20and%20Research\Raw%20Data%20and%20Information\Federal%20Government%20Sources\Canadian%20International%20Merchandise%20Trade\Annual%22) folder and write a csv file file for data visualization.

```{r Bind_ExportAnnual}

tic()

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")

# Initialize an empty list to store data frames
export_list <- list()

# Loop through the years 2013 to 2023

for (year in 2003:2023) {
  # Construct the file name
  file_name <- paste0("Export", year, ".csv")
  
  # Read the CSV file
  Exportyear <- fread(file_name, colClasses = c(HS8_Code = "character"))
  
  # Add the data frame to the list
  export_list[[length(export_list) + 1]] <- Exportyear
}

# Bind all data frames in the list into one data frame
ExportFULL <- bind_rows(export_list, .id = NULL)

toc()
```

```{r Write_ExportFULL}

tic()

# Create the ImportFULL dataframe with Date, Year, and Month columns
ExportFULL <- ExportFULL %>%
  mutate(
    Year = substr(Date, 1, 4),       # Extract the first 4 characters as Year
    Month = substr(Date, nchar(Date)-1, nchar(Date))  # Extract the last 2 characters as Month
  )

ExportFULL$Date <- ymd(paste0(ExportFULL$Year,ExportFULL$Month,"01", sep = "-"))

#drop year and month columns
ExportFULL <- ExportFULL[, -c(9, 10)]

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")

fwrite(ExportFULL, "ExportFULL.csv")

toc()
```

## 2. Imports: Load and Bind All Years

The code chunk under this heading is an example of how to append all annual ***import*** data frames found in the [Annual](%22\sfp.idir.bcgov\S104\S04004\Statistics%20and%20Research\Raw%20Data%20and%20Information\Federal%20Government%20Sources\Canadian%20International%20Merchandise%20Trade\Annual%22) folder and write a csv file file for data visualization.

```{r Bind_ImportAnnual}
tic()  

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade/Annual")


# Initialize an empty list to store data frames
import_list <- list()  

# Loop through the years 1988 to 2023 
for (year in 2003:2023) 
  {   
  # Construct the file name   
  file_name <- paste0("Import", year, ".csv")      
  # Read the CSV file   
  Importyear <- fread(file_name, colClasses = c(HS10_Code = "character"))      
  # Add the data frame to the list   
  import_list[[length(import_list) + 1]] <- Importyear }  

# Bind all data frames in the list into one data frame 
ImportFULL <- bind_rows(import_list, .id = NULL)  

toc()
```

```{r Write_ImportFULL}
tic() 

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")

# Create the ImportFULL dataframe with Date, Year, and Month columns 
ImportFULL <- ImportFULL %>%
  mutate(Year = substr(Date, 1, 4),  
         
# Extract the first 4 characters as Year
    Month = substr(Date, nchar(Date)-1, nchar(Date))) 

# Extract the last 2 characters as Month
    
ImportFULL$Date <- ymd(paste0(ImportFULL$Year,ImportFULL$Month,"01", sep = "-")) 

#drop year and month columns 

ImportFULL <- ImportFULL[, -c(9, 10)]  

fwrite(ImportFULL, "ImportFULL.csv")  
                                           
toc() 
```

## 3. Trade: Full Join Export FULL and Import FULL

Need to merge both and calculate trade balance

```{r Read_ExpImpFULL}

# Run if you do not have ExportFULL and ImportFULL in the environment

tic()

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")

ExportFULL <- fread("ExportFULL.csv", colClasses = c(HS8_Code = "character"))
ImportFULL <- fread("ImportFULL.csv", colClasses = c(HS10_Code = "character"))

toc()
```

```{r Clean_ImportFULL}

tic()

# Extract the first 8 characters as HS8 for Imports
ImportFULL_HS8 <- ImportFULL %>%
  mutate(
    HS8_Code = substr(HS10_Code, 1, 8))       

# Remove HS10_Code
ImportFULL_HS8 <- ImportFULL_HS8[, -c(1)]

# Group by HS8_Code
ImportFULL_HS8 <- ImportFULL_HS8 %>% 
  group_by(HS8_Code, 
           Date, 
           Country, 
           Province, 
           State, 
           UOM) %>% 
  summarise(Value = sum(Value),
            Quantity = sum(Quantity))


# Format as Date

ImportFULL_HS8$Date <- as.Date(ImportFULL_HS8$Date)
ExportFULL$Date <- as.Date(ExportFULL$Date)

toc()
```

```{r Join_TradeFULL}

tic()

TradeFULL <- full_join(ExportFULL, ImportFULL_HS8,
                       by = c("HS8_Code",
                              "Date",
                              "Country",
                              "Province",
                              "State",
                              "UOM"),
                       copy = FALSE,
                       suffix = c("_X","_M"),
                       keep = NULL)

#replace NAs with 0

TradeFULL[is.na(TradeFULL)] <- 0

# calculate trade balance

TradeFULL$Value_Balance <- TradeFULL$Value_X-TradeFULL$Value_M

TradeFULL$Quantity_Balance <- TradeFULL$Quantity_X-TradeFULL$Quantity_M

# Re-arrange the columns in a useful order
TradeFULL <- TradeFULL[,c(1:5,8,6:7,9:12)]


TradeFULL <- pivot_longer(TradeFULL,
                          7:12,
                          names_to = "Estimate",
                          values_to = "Value")

toc()
```

```{r write_TradeFULL}

tic()

setwd("//sfp.idir.bcgov/S104/S04004/Statistics and Research/Raw Data and Information/Federal Government Sources/Canadian International Merchandise Trade")


fwrite(TradeFULL, "TradeFULL.csv",
       row.names = FALSE,
       col.names = TRUE,
       buffMB = 10)

toc()
```
