---
title: "BC AF Trade Tool Update"
author: "Leila Bautista"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Canadian International Merchandise Trade Database (CIMTD)

This RMarkdown provides chunks of R code to load, filter and combine data from Open Government [website](https://search.open.canada.ca/opendata/?sort=metadata_modified+desc&search_text=cimt&page=1). The trade data is used by the Ministry of Agriculture and Food's Sector Insights and Corporate Initiatives (SICI) Unit to produce annual Sector Tables, Trade Dashboards and other information products.

Contact [alstats\@gov.bc.ca](mailto:alstats@gov.bc.ca) for any questions.

# Set-up & Library

```{r Set-up, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory

knitr::opts_knit$set(root.dir = "C:/Users/lebautis/Government of BC/SICI Data Library Subsite - Unpublished Data")
```

```{r Library}

# Load library
library(pacman)
p_load(dplyr,
       data.table,
       tidyr,
       lubridate,
       tictoc,
       splitstackshape,
       stringr)
```

# Load and read latest data

Export data

```{r Export Latest}
tic()

# Set the working directory to desktop
setwd("C:/Users/lebautis/Desktop/")

# Step 1: Create a temporary file
temp_file <- tempfile()

# Step 2: Fetch the online file using download.file()
# Replace 'url_of_the_file' with the actual URL of the file you want to download
url <- "https://www150.statcan.gc.ca/n1/pub/71-607-x/2021004/zip/CIMT-CICM_Dom_Exp_2024.zip"
download.file(url, temp_file)

# Step 3: Extract the file from the temp file using unz() command
# Replace 'file_inside_zip' with the actual name of the file inside the ZIP archive
file_inside_zip <- "CIMT-CICM_Dom_Exp_2024/ODPFN016_202405N.csv"
Export <- read.csv(unz(temp_file, file_inside_zip))

# Step 4: Remove the temp file using unlink()
unlink(temp_file)

# Set the working directory to SharePoint
setwd("C:/Users/lebautis/Government of BC/SICI Data Library Subsite - Unpublished Data")

# Load the list of agriculture, seafood and food manufacturing HS codes 
Commodity_REF <- fread("Commodity_Classification.csv",
                       colClasses = c(HS8_Code = "character"))

# Read the CSV file into a data frame
Export_Update <- read.csv(file = "ODPFN016_202412N.csv", stringsAsFactors = FALSE, colClasses = c(HS8 = "character", Value = "numeric", Quantity = "numeric"))


# Rename variables then join left with the Commodity_REF to fileter for relevant HS codes only
Export2024 <- Export_Update %>% 
  rename (Date = `YearMonth.AnnéeMois`,
          HS8_code = `HS8`,
          Country = `Country.Pays`,
          State = `State.État`,
          Value = `Value.Valeur`,
          Quantity = `Quantity.Quantité`,
          UOM = `Unit.of.Measure.Unité.de.Mesure`
  ) %>%
  merge(Commodity_REF,by="HS8_code") %>% 
  select(1:8)

# reformat the date column
Export_Update$Date <- ym(Export_Update$Date)


# to write code: bind with Trade FULL


toc()
```

Import data

```{r Import Latest}

tic()


# Set the working directory to desktop
setwd("C:/Users/lebautis/Desktop/")

# Step 1: Create a temporary file
temp_file <- tempfile()

# Step 2: Fetch the online file using download.file()
# Replace 'url_of_the_file' with the actual URL of the file you want to download
url <- "https://www150.statcan.gc.ca/n1/pub/71-607-x/2021004/zip/CIMT-CICM_Dom_Exp_2024.zip"
download.file(url, temp_file)

# Step 3: Extract the file from the temp file using unz() command
# Replace 'file_inside_zip' with the actual name of the file inside the ZIP archive
file_inside_zip <- "CIMT-CICM_Dom_Exp_2024/ODPFN016_202405N.csv"
Import <- read.csv(unz(temp_file, file_inside_zip))

# Step 4: Remove the temp file using unlink()
unlink(temp_file)

# Read the CSV file into a data frame
Import_Update <- read.csv("ODPFN014_202412N.csv", stringsAsFactors = FALSE, colClasses = c(HS10 = "character", Value = "numeric", Quantity = "numeric"))

# Set the working directory to SharePoint
setwd("C:/Users/lebautis/Government of BC/SICI Data Library Subsite - Unpublished Data")


# Load the list of agriculture, seafood and food manufacturing HS codes 
Commodity_REF <- read.csv("HS8X_2022_AAFC_REF.csv", colClasses = "character")
Commodity_REF <- Commodity_REF %>% rename (HS8_code = HS8_Code)
# Rename variables
Import_Update <- Import_Update %>% 
  rename (Date = `YearMonth.AnnéeMois`,
          HS10_code = `HS10`,
          Country = `Country.Pays`,
          State = `State.État`,
          Value = `Value.Valeur`,
          Quantity = `Quantity.Quantité`,
          UOM = `Unit.of.Measure.Unité.de.Mesure`
  ) %>%
  select(1:8)

# reformat the date column
Import_Update$Date <- ym(Import_Update$Date)

# Replace blank, NA, or NULL values in 'State' with 'Country'
Import_Update$State[is.na(Import_Update$State) | Import_Update$State == ""] <- "Country"

# Transform HS10 to HS8 by summing Value and Quantity then join left with the Commodity_REF to fileter for relevant HS codes only
Import_Update <- Import_Update %>%
  mutate(HS8_code = substr(HS10_code, 1, 8))  # Trim first 8 characters 
  
## create new column
Import_Update <- Import_Update %>%
  select(-HS10_code) %>%  # Remove original HS10 column
  group_by(Date, HS8_code, Country, State, Province, UOM) %>%
  summarise(
    Value = sum(Value, na.rm = TRUE),
    Quantity = sum(Quantity, na.rm = TRUE),
    .groups = "drop")

# to write code: bind with Trade FULL
Import_Update <- Import_Update %>% merge(Commodity_REF,by="HS8_code")

toc()

```

# Calculate the trade balance

```{r Calculate Trade Balance Update}
tic()

Trade_Update <- full_join(Export_Update, Import_Update,
                       by = c("HS8_code",
                              "Date",
                              "Country",
                              "Province",
                              "State",
                              "UOM"),
                       copy = FALSE,
                       suffix = c("_X", "_M"),
                       keep = NULL)



#replace NAs with 0

Trade_Update[is.na(Trade_Update)] <- 0

# calculate trade balance

Trade_Update$Value_Balance <- Trade_Update$Value_X-Trade_Update$Value_M

Trade_Update$Quantity_Balance <- Trade_Update$Quantity_X-Trade_Update$Quantity_M

# Re-arrange the columns in a useful order
Trade_Update <- Trade_Update[,c(1:5,8,6:7,9:12)]


Trade_Update <- pivot_longer(Trade_Update,
                          7:12,
                          names_to = "Estimate",
                          values_to = "Value")

toc()
```

# Append to TradeFULL

```{r APpend to TradeFULL}

# Set the working directory to the location of the CSV file
setwd("C:/Users/lebautis/Desktop/CIMTD")

# Read the CSV file into a data frame
TradeFULL <- read.csv("TradeFULL.csv", stringsAsFactors = FALSE, colClasses = c(HS8_Code = "character", Value = "numeric"))
TradeFULL <- TradeFULL %>% rename (HS8_code = HS8_Code)

TradeFULL$Date <- ymd(TradeFULL$Date)

# Ensure both data frames have the same columns
if(all(colnames(TradeFULL) == colnames(Trade_Update))) {
  # Combine the two data frames row-wise
  CombinedTradeData <- rbind(TradeFULL, Trade_Update)
  # Print the combined data
  print(CombinedTradeData)
} else {
  # If columns do not match, give an error message
  print("The columns of the two data frames do not match.")
}


```

# Write CSV

For now, upload to sharepoint manually until we figure out M365R function

```{r write CSV}

tic()

# Set the working directory to the location of the CSV file
setwd("C:/Users/lebautis/Desktop/CIMTD")

fwrite(CombinedTradeData, "TradeFULL.csv",
       row.names = FALSE,
       col.names = TRUE,
       buffMB = 10)

toc()
```
